---
title: "Microbial Community Analyses"
author: "Aeran Coughlin"
date: "2026-05-11"
output:
  pdf_document: default
  html_document: default
---

This R Markdown file contains the fungal community analyses employed for the manuscript "Microbial assemblages of coastal forested wetland soil promote growth of expanding woody shrub" by Coughlin & Wright. 

### Load Packages and Data Files
```{r}
library(tidyverse)
library(car)
library(vegan)
library(pals)
library(dbstats)
library(ape)
library(FUNGuildR)
library(betareg)
library(VennDiagram)
library(randomForest)
library(phyloseq)
library(corrr)
library(patchwork)
library(ggpubr)
library(lmtest)
library(lsmeans)
library(multcomp)

fun.otu <- read.csv("fun.table.csv")
fun.tax <- read.csv("fun.tax.csv")

pots <- read.csv("PSF-exp-2022_pot-data.csv")
pp <- read.csv("PSF-exp-2022_plant-performance.csv")
myco <- read.csv("PSF-exp-2022_myco.csv")
```

### Data Cleaning

Filter only greenhouse data.

```{r}
# select only greenhouse samples 
fun.otu.gh <- fun.otu %>% 
  subset(Type=="GH") %>% 
  arrange(plot.code) %>% 
  dplyr::select(-c(2)) %>%
  rename("pot.code"=plot.code)
```

Remove singletons.

```{r}
# convert to presence / absence
fun.otu.pa <- decostand(fun.otu.gh %>% 
                          arrange(pot.code) %>% 
                          dplyr::select(-c(1)), method = "pa") 

# select only ASVs with more than one occurrence (exclude singletons)
fun.otu.nonsing <- fun.otu.pa %>% 
  dplyr::select(where(~ is.numeric(.) && sum(., na.rm = TRUE) > 1)) %>% 
  colnames()

# extract only non-singletons from original otu table
fun.otu.matrix <- fun.otu.gh %>% 
  dplyr::select(all_of(fun.otu.nonsing))
  
# add pot code column 
fun.otu.labeled <- cbind(fun.otu.gh %>% dplyr::select(pot.code), fun.otu.matrix)

fun.otu.pots <- fun.otu.gh %>% 
  dplyr::select(1) %>% 
  left_join(pots, by="pot.code") %>% 
  dplyr::select(pot.code, species, inoculum) %>% 
  mutate(species = ifelse(species == "MYCE", "MOCE", species))

fun.otu.list <- fun.otu.matrix %>% 
  pivot_longer(names_to="OTU", values_to="count", 1:1113) %>% 
  dplyr::select(OTU) %>% 
  unique()
```

Remove sterile treatment ASVs from treatment samples to use for certain analyses

```{r}
# identify ASVs that are found in sterile treatments
fun.otu.ster <- fun.otu.pots %>% 
  filter(inoculum=="Sterile") %>% 
  dplyr::select(pot.code) %>% 
  left_join(fun.otu.labeled, by="pot.code") %>% 
  dplyr::select(-c(1)) %>% 
  dplyr::select(where(~ is.numeric(.) && sum(., na.rm = TRUE) > 0)) %>% 
  colnames() # 373 ASVs

# phylum breakdown in sterile treatment
fun.ster.sum <- fun.otu.labeled %>% 
  dplyr::select(all_of(fun.otu.ster))

fun.ster.sum2 <-  fun.otu.pots %>% 
  dplyr::select(pot.code, inoculum) %>%
  cbind(fun.ster.sum) %>% 
  filter(inoculum=="Sterile") %>% 
  dplyr::select(-c(inoculum)) %>% 
  pivot_longer(names_to="OTU", values_to="count", 2:374) %>% 
  group_by(OTU) %>% 
  summarise(sum=sum())

# create table of ASVs only found in inoculated treatments
# (exclude those found in sterile)
fun.otu.inoc <- fun.otu.labeled %>% 
  dplyr::select(-all_of(fun.otu.ster)) %>% 
  left_join(fun.otu.pots %>% dplyr::select(pot.code, inoculum), by="pot.code") %>%
  filter(inoculum!="Sterile") %>% 
  dplyr::select(-c(inoculum)) # 741 ASVs

#fun.tax.inoc <- fun.otu.inoc %>% 
#  pivot_longer() %>% 
#  left_join(fun.tax)
```



Quantify sampling depth across pots for assessing sampling evenness and rarefaction.

```{r}
# calculate number of sequence reads per pot
pot.read.n <- fun.otu.labeled %>% 
  mutate(rowsum = rowSums(select_if(., is.numeric), na.rm = TRUE)) %>% 
  dplyr::select(pot.code, rowsum) # smallest size: 32762 reads

# create histogram of sequence reads across pots
ggplot(pot.read.n) + 
  geom_histogram(aes(rowsum), bins=10, color="black", fill="pink") + 
  xlab("Sequence reads per pot") + 
  ylab("n") + 
  theme_classic()
```

Identify the 10% most abundant ASVs for manual BLAST calling

```{r}
# select only top 10% most abundant ASVs from dataset
fun.otu.total <- fun.otu.matrix %>% 
  colSums() %>% 
  as.data.frame() %>% 
  rownames_to_column(var="OTU") %>% 
  rename("count"=".") %>% 
  arrange(desc(count)) %>% 
  top_n(111, count) %>% 
  arrange(OTU) %>% 
  left_join(fun.tax, by="OTU")

# export for manual calling / confirmation
write.csv(fun.otu.total, "fun.otu.total.csv")

# reads in the top 10% most abundant ASVs
top.10 <- fun.otu.total %>% summarise(sum=sum(count)) 

# total reads in dataset 
all.otus <- fun.otu.matrix %>% 
  colSums() %>% 
  as.data.frame() %>% 
  rownames_to_column(var="OTU") %>%
  summarise(sum=sum(.)) 

# proportion of reads that 10% encompasses
(top.10 / all.otus) * 100

# import new file post-manual correction
fun.tax.cor <- read.csv("fun.otu.total_manually-corrected.csv")
```

Replace corrected taxonomic assignments in taxonomy table

```{r}
# create df of 90% least abundant ASVs (not manually checked)
fun.tax.uncor <- fun.tax %>% 
  filter(OTU %in% fun.otu.list$OTU) %>% 
  filter(! OTU %in% fun.tax.cor$OTU)

# bind with top 10% of ASVs, which have now been manually checked
fun.tax.new <- fun.tax.cor %>% 
  dplyr::select(-c(1,3,12:14)) %>% 
  rbind(fun.tax.uncor)
```


### Composition and Diversity Analyses

Diversity 

```{r}
# create data frame for diversity variables
fun.otu.div <- fun.otu.pots %>% 
  left_join(pp %>% dplyr::select(pot.code, bm.rate, psf), by="pot.code")

# richness
fun.otu.div$rich <- specnumber(fun.otu.labeled)

# rarefied richness
fun.otu.div$rare <- rarefy(fun.otu.matrix, 32762)

# shannon
fun.otu.div$shannon <- diversity(fun.otu.matrix, index = "shannon")

# evenness 
fun.otu.div$even <- fun.otu.div$shannon / log(fun.otu.div$rich)

# relationship between raw richness and rarefied richness 
cor(fun.otu.div$rich, fun.otu.div$rare) # R2 = 0.9989917

ggplot(fun.otu.div) + 
  geom_abline(slope=1, intercept=0) + 
  geom_point(aes(x=rich, y=rare), shape=21, color="black", fill="pink", size=3) + 
  theme_classic() + 
  ylab("Rarefied Richness") + 
  xlab("Richness")

# analyze differences among treatment groups
rich.lm <- lm(rich ~ species * inoculum, data=fun.otu.div)
Anova(rich.lm)

even.lm <- lm(even ~ species * inoculum, data=fun.otu.div)
Anova(even.lm)

# analyze relationships with plant performance
rich.psf <- lm(psf ~ rich, data=fun.otu.div)
Anova(rich.psf)

ggplot(fun.otu.div) + 
  geom_point(aes(x=rich, y=psf), size=3) + 
  geom_smooth(aes(x=rich, y=psf), method="lm") + 
  theme_classic() + 
  ylab("Plant Performance") + 
  xlab("Fungal Richness")
```

Multidimensional Scaling Analysis

```{r}
# calculate relative abundance 
fun.matrix.rel <- decostand(fun.otu.matrix, method = "total")

# calculate dissimilarity matrix 
fun.matrix.dist <- vegdist(fun.matrix.rel, method = "bray")

# run mds 
fun.gh.mds <- metaMDS(fun.matrix.dist, distance = "bray", trymax=1000)

# goodness of fit
stressplot(fun.gh.mds)

# extract MDS values for visualization
fun.scores <- as.data.frame(vegan::scores(fun.gh.mds))

fun.scores$pot.code = fun.otu.pots$pot.code

fun.scores.meta <- fun.scores %>% 
  left_join(fun.otu.pots, by="pot.code") %>% 
  left_join(pp %>% dplyr::select(pot.code, psf, bm.rate), by="pot.code")
```

Then, without 9F, which we are missing biomass data for, to test how plant performance fits onto composition ordination.

```{r}
# calculate relative abundance 
fun.matrix.rel.psf <- decostand(fun.otu.gh %>% 
                                  arrange(pot.code) %>% 
                                  filter(pot.code!="9F") %>% 
                                  dplyr::select(-c(1)), 
                                method = "total")

fun.otu.rich.psf <- fun.otu.div %>% arrange(pot.code) %>% filter(pot.code!="9F")

# calculate dissimilarity matrix
fun.matrix.dist.psf <- vegdist(fun.matrix.rel.psf, method = "bray")

# run mds 
fun.gh.mds.psf <- metaMDS(fun.matrix.dist.psf, distance = "bray", trymax=1000)

# fit relative performance data to mds
psf.envfit <- envfit(fun.matrix.dist.psf, fun.otu.rich.psf$psf, permutations = 999, na.rm = TRUE)

psf.envfit
```

Fungal guild

```{r}
# format df for fun guild assignments
fun.tax.gh <- fun.tax.new %>% 
  unite(col = "Taxonomy", 
        c(Kingdom, Phylum, Class, Order, Family, Genus, Species), 
        sep = "; ", remove = TRUE) %>% 
  dplyr::select(OTU, Taxonomy)

# assign guild / trophic mode
fun.guild.gh <- funguild_assign(fun.tax.gh, db = get_funguild_db(), tax_col = "Taxonomy")

# clean guild df
fun.guild.plot <- fun.otu.labeled %>% 
  arrange(pot.code) %>% 
  gather("OTU", "count", 2:1114) %>% 
  left_join(fun.guild.gh, by="OTU") %>% 
  dplyr::select(pot.code, OTU,Taxonomy, trophicMode, count) %>%
  mutate(trophicMode = replace_na(trophicMode, "Unknown")) %>% 
  mutate(trophicMode = replace(trophicMode, trophicMode== " Saprotroph", "Saprotroph")) %>% 
  mutate(trophicMode = replace(trophicMode, trophicMode== " Pathotroph-Pathotroph-Saprotroph", "Pathotroph-Pathotroph-Saprotroph")) %>% 
  mutate(trophicMode = replace(trophicMode, trophicMode== " Saprotroph-Saprotroph-Symbiotroph", "Saprotroph-Saprotroph-Symbiotroph")) %>%   group_by(pot.code, trophicMode) %>%
  summarise(count.sum=sum(count)) %>% 
  left_join(fun.otu.pots, by="pot.code")

# count total reads per pot
fun.guild.sum <- fun.otu.labeled %>% 
  arrange(pot.code) %>% 
  gather("OTU", "count", 2:1114) %>% 
  left_join(fun.guild.gh, by="OTU") %>% 
  dplyr::select(pot.code, OTU,Taxonomy, trophicMode, count) %>%
  group_by(pot.code) %>%
  summarise(total.count=sum(count))

# calculate relative abundance of each trophic mode in each pot
fun.guild.rel <- pp %>% 
  dplyr::select(pot.code, bm.rate, psf, growth.time) %>% 
  left_join(fun.guild.plot, by="pot.code") %>% 
  left_join(fun.guild.sum, by="pot.code") %>% 
  mutate(rel.count = count.sum/total.count)

# analyses 

# extract specific trophic modes
# saprotrophs
sap.rel <- fun.guild.rel %>% 
  filter(trophicMode=="Saprotroph") %>% 
  remove_missing()

# symbiotrophs
sym.rel <- fun.guild.rel %>% 
  filter(trophicMode=="Symbiotroph") %>% 
  remove_missing()

# pathotrophs
path.rel <- fun.guild.rel %>% 
  filter(trophicMode=="Pathotroph") %>% 
  remove_missing()

# extract specific guilds 
# plant pathogens
fun.guild.plantpath <- fun.otu.labeled %>%
  gather("OTU", "count", 2:1114) %>% 
  left_join(fun.guild.gh, by="OTU") %>% 
  dplyr::select(pot.code, OTU,Taxonomy, guild, count) %>%
  group_by(pot.code, guild) %>%
  summarise(count.sum=sum(count)) %>% 
  filter(guild=="|Plant Pathogen|") %>% 
  group_by(pot.code) %>%
  summarise(plant.path=sum(count.sum)) %>% 
  left_join(fun.guild.sum, by="pot.code") %>% 
  left_join(pp, by="pot.code") %>% 
  mutate(rel.count = plant.path/total.count)

# fungal parasites
fun.guild.funpar <- fun.otu.labeled %>%
  gather("OTU", "count", 2:1114) %>% 
  left_join(fun.guild.gh, by="OTU") %>% 
  dplyr::select(pot.code, OTU,Taxonomy, guild, count) %>%
  group_by(pot.code, guild) %>%
  summarise(count.sum=sum(count)) %>% 
  filter(guild=="|Fungal Parasite|") %>% 
  group_by(pot.code) %>%
  summarise(fun.par=sum(count.sum)) %>% 
  left_join(fun.guild.sum, by="pot.code") %>% 
  left_join(pp, by="pot.code") %>% 
  mutate(rel.count = fun.par/total.count)

# ectomycorrhizae
fun.guild.ecm <- fun.otu.labeled %>%
  gather("OTU", "count", 2:1114) %>% 
  left_join(fun.guild.gh, by="OTU") %>% 
  dplyr::select(pot.code, OTU,Taxonomy, guild, count) %>%
  group_by(pot.code, guild) %>%
  summarise(count.sum=sum(count)) %>% 
  filter(guild=="Ectomycorrhizal" | guild=="|Ectomycorrhizal|") %>% 
  group_by(pot.code) %>%
  summarise(ecm=sum(count.sum)) %>% 
  left_join(fun.guild.sum, by="pot.code") %>% 
  left_join(pp, by="pot.code") %>% 
  filter(species=="PITA") %>% 
  mutate(rel.count = ecm/total.count)

# specific ectomycorrhizal genera
fun.ecm.gen <- fun.otu.labeled %>%
  gather("OTU", "count", 2:1114) %>% 
  left_join(fun.guild.gh, by="OTU") %>% 
  dplyr::select(pot.code, OTU, guild, count) %>%
  filter(guild=="Ectomycorrhizal" | guild=="|Ectomycorrhizal|") %>% 
  left_join(fun.tax, by="OTU") %>% 
  dplyr::select(pot.code, Family, Genus, count) %>% 
  group_by(pot.code, Family, Genus) %>%
  summarise(gen.sum=sum(count)) %>% 
  left_join(pp, by="pot.code")

# count number of unique ectomycorrhizal ASVs
fun.otu.labeled %>%
  gather("OTU", "count", 2:1114) %>% 
  left_join(fun.guild.gh, by="OTU") %>% 
  dplyr::select(pot.code, OTU, guild, count) %>%
  filter(guild=="Ectomycorrhizal" | guild=="|Ectomycorrhizal|") %>% 
  dplyr::select(OTU) %>% 
  unique() %>% 
  count() 

# specific ectomycorrhizal genera by treatment
fun.ecm.treat <- fun.ecm.gen %>% 
  group_by(inoculum, Genus) %>% 
  summarize(sum=sum(gen.sum))

# arbuscular mycorrhizae
fun.guild.am <- fun.otu.gh %>%
  gather("OTU", "count", 2:8679) %>% 
  left_join(fun.guild.gh, by="OTU") %>% 
  dplyr::select(pot.code, OTU,Taxonomy, guild, count) %>%
  group_by(pot.code, guild) %>%
  summarise(count.sum=sum(count)) %>% 
  filter(str_detect(guild, "Arbusc")) %>%
  group_by(pot.code) %>%
  summarise(am=sum(count.sum)) %>% 
  left_join(fun.guild.sum, by="pot.code") %>% 
  left_join(pp, by="pot.code") %>% 
  mutate(rel.count = am/total.count)

# count unique arbuscular mycorrhizal ASVs
fun.otu.gh %>%
  gather("OTU", "count", 2:8679) %>% 
  left_join(fun.guild.gh, by="OTU") %>% 
  dplyr::select(pot.code, OTU,Taxonomy, guild, count) %>%
  filter(str_detect(guild, "Arbusc")) %>% 
  dplyr::select(OTU) %>% 
  unique() %>% 
  count()
  
# dark septate endophytes
fun.dse <- fun.otu.labeled %>%
  gather("OTU", "count", 2:1114) %>% 
  left_join(fun.otu.pots %>% dplyr::select(pot.code, species, inoculum), by="pot.code") %>% 
  left_join(fun.guild.gh, by="OTU") %>% 
  dplyr::select(pot.code, species, inoculum, OTU, Taxonomy, guild, count, trophicMode, notes) %>%
  filter(grepl('Dark Septate', notes)) %>% 
  filter(count!=0)

# dse abundance by pot
fun.dse.count <- fun.dse %>% 
  group_by(pot.code) %>% 
  dplyr::select(pot.code, count) %>% 
  summarise(count=sum(count)) %>% 
  filter(pot.code!="6A") %>% 
  left_join(fun.guild.sum, by="pot.code") %>% 
  left_join(pp, by="pot.code") %>% 
  mutate(rel.count = count/total.count)

# count unique dse ASVs
fun.dse %>% 
dplyr::select(OTU) %>% 
  unique() %>% 
  count() 

# build models 

# symbiotrophs
# performance and relative abundance 
sym.psf.lm <- lm(psf~rel.count * species, data=sym.rel)
summary(sym.psf.lm)

# relative abundance and treatment group
waldtest(betareg(rel.count ~ species, data=sym.rel))
waldtest(betareg(rel.count ~ inoculum, data=sym.rel))
waldtest(betareg(rel.count ~ species*inoculum, data=sym.rel))

sym.br <- betareg(rel.count ~ species*inoculum, data=sym.rel)
Anova(betareg(rel.count ~ species*inoculum, data=sym.rel)) # wald chi square

# post hoc test
sym.rg <- ref.grid(sym.br)        
sym.lsm = lsmeans(sym.rg, ~ species*inoculum)
cld(sym.lsm, alpha =  0.05)

# ectomycorrhizae 
# performance and relative abundance 
ecm.psf.lm <- lm(psf~rel.count, data=fun.guild.ecm)
summary(ecm.psf.lm)

# relative abundance and treatment group
waldtest(betareg(rel.count ~ inoculum, data=fun.guild.ecm))
ecm.br <- betareg(rel.count ~ inoculum, data=fun.guild.ecm)

# post hoc test
ecm.rg <- ref.grid(ecm.br)        
ecm.lsm = lsmeans(ecm.rg, ~ inoculum)
cld(ecm.lsm, alpha =  0.05)

# arbuscular mycorrhizae
# performance and relative abundance 
am.psf.lm <- lm(psf~rel.count, data=fun.guild.am %>% filter(species=="ILGL"))
summary(am.psf.lm)

# relative abundance and treatment group
# beta regression does not work for this model, using LM
am.count <- lm(rel.count ~ inoculum, data=fun.guild.am %>% filter(species=="ILGL"))
Anova(am.count)

# post-hoc test
am.count.aov <- aov(am.count)
TukeyHSD(am.count.aov, conf.level=0.95)

# dark septate endophytes
# performance and relative abundance 
dse.psf.lm <- lm(psf~rel.count, data=fun.dse.count)
summary(dse.psf.lm)

# relative abundance and treatment group
waldtest(betareg(rel.count ~ species, data=fun.dse.count))
waldtest(betareg(rel.count ~ inoculum, data=fun.dse.count))
waldtest(betareg(rel.count ~ species*inoculum, data=fun.dse.count))

dse.br <- betareg(rel.count ~ species*inoculum, data=fun.dse.count)
Anova(betareg(rel.count ~ species*inoculum, data=fun.dse.count)) # wald chi square

# post hoc test
dse.rg <- ref.grid(dse.br)        
dse.lsm = lsmeans(dse.rg, ~ species*inoculum)
cld(dse.lsm, alpha =  0.05)

# total pathotrophs
# performance and relative abundance 
path.psf.lm <- lm(psf~rel.count, data=path.rel)
summary(path.psf.lm)

# relative abundance and treatment group
waldtest(betareg(rel.count ~ species, data=path.rel))
waldtest(betareg(rel.count ~ inoculum, data=path.rel))
waldtest(betareg(rel.count ~ species*inoculum, data=path.rel))

path.br <- betareg(rel.count ~ species * inoculum, data=path.rel)
Anova(betareg(rel.count ~ species*inoculum, data=path.rel)) # wald chi square

# post hoc test
path.rg <- ref.grid(path.br)        
path.lsm = lsmeans(path.rg, ~ species*inoculum)
cld(path.lsm, alpha =  0.05)

# plant pathogens
# performance and relative abundance 
plantpath.psf.lm <- lm(psf~rel.count, data=fun.guild.plantpath)
summary(plantpath.psf.lm)

# relative abundance and treatment group
plantpath.count <- lm(rel.count ~ species*inoculum, data=fun.guild.plantpath)
summary(plantpath.count)

# fungal parasites
# performance and relative abundance 
funpar.psf.lm <- lm(psf~rel.count, data=fun.guild.funpar)
summary(funpar.psf.lm)

# relative abundance and treatment group
funpar.count <- lm(rel.count ~ species*inoculum, data=fun.guild.funpar)
summary(funpar.count)

# saprotrophs
# performance and relative abundance 
sap.psf.lm <- lm(psf~rel.count, data=sap.rel)
summary(sap.psf.lm)

# relative abundance and treatment group
waldtest(betareg(rel.count ~ species, data=sap.rel))
waldtest(betareg(rel.count ~ inoculum, data=sap.rel))
waldtest(betareg(rel.count ~ species*inoculum, data=sap.rel))

sap.br <- betareg(rel.count ~ species*inoculum, data=sap.rel)
Anova(betareg(rel.count ~ species*inoculum, data=sap.rel)) # wald chi square

# post hoc test
sap.rg <- ref.grid(sap.br)        
sap.lsm = lsmeans(sap.rg, ~ species*inoculum)
cld(sap.lsm, alpha =  0.05)
```

### Create figure of composition, diversity, and trophic mode abundance (Figure 3).

```{r}
# create color palette
psf.bw <- c("Forest"="gray30", "Shrub"="gray", "Sterile"="gray95")

# create guild data frame that includes dse assignments
troph.fig <- fun.guild.rel %>% 
  filter(trophicMode=="Saprotroph" | 
           trophicMode=="Pathotroph" |
           trophicMode=="Symbiotroph") %>% 
  dplyr::select(pot.code, species, inoculum, trophicMode, rel.count) %>% 
  rbind(fun.dse.count %>% 
          mutate(trophicMode = "DSE") %>% 
          dplyr::select(pot.code, species, inoculum, trophicMode, rel.count))

# nmds 
fun.mds <- ggplot(fun.scores.meta, aes(x=NMDS1, y=NMDS2)) +
  stat_ellipse(aes(fill=inoculum), geom = "polygon", color="black", alpha=0.5) + 
  geom_point(aes(shape=species, fill=inoculum), color="black", size=3) + 
  theme_classic() + 
  scale_shape_manual(values=c(21,24,22)) + 
  scale_fill_manual(values=psf.bw) + 
  xlab("NMDS 1") + 
  ylab("NMDS 2") + 
  guides(fill = guide_legend(override.aes = list(shape=21)))

# richness across species X inocula
fun.rich <- ggplot(fun.otu.div) + 
  geom_boxplot(aes(y=rich, x=species, fill=inoculum)) + 
  theme_classic() + 
  scale_fill_manual(values=psf.bw) + 
  ylab("Richness")

# evenness across species X inocula
fun.even <- ggplot(fun.otu.div) + 
  geom_boxplot(aes(y=even, x=species, fill=inoculum)) + 
  theme_classic() + 
  scale_fill_manual(values=psf.bw) + 
  ylab("Evenness")

# relative abundance of trophic modes across species x inocula
troph <- ggplot(troph.fig %>% 
                  mutate(rel.count = replace(rel.count, 
                                             trophicMode=="Pathotroph" & rel.count>0.4, NA)), 
                aes(x=inoculum, y=rel.count)) + 
  geom_boxplot(aes(fill=inoculum)) +
  theme_classic() + 
  xlab("Inoculum") + 
  ylab("Relative Abundance") + 
  scale_fill_manual(values=psf.bw) + 
  theme(legend.position="NONE", 
        panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank()) + 
  facet_grid(rows=vars(trophicMode), cols=vars(species), scale="free_y")

# combine composition and richness plots for top row
comp.rich <- ggarrange(fun.mds, fun.rich, fun.even,
                       nrow=1, ncol=3, common.legend=TRUE, labels=c("A","B","C"))

# add trophic mode plot below
comp.rich.troph <- ggarrange(comp.rich, troph, 
                             nrow=2, ncol=1, heights=c(1,1.5), labels=c("","D"))
comp.rich.troph 

# save
ggsave("comp.rich.troph.png", dpi=600, width=7, height=7, plot=comp.rich.troph)

```


### Visualize Fungal Phyla (with and without Sterile ASVs)

First, build data frames for each species to summarize the relative abundance of each Phylum, including sterile ASVs.

```{r}
# create data frame of treatment data and ASV abundance
fun.otu.abun <- fun.otu.labeled %>%
  gather(OTU, abun, 2:1113) %>% 
  left_join(fun.tax.new, by="OTU") %>%
  left_join(fun.otu.pots, by="pot.code") %>%
  dplyr::select(pot.code, species, inoculum, OTU, abun)

fun.phy.sum <- fun.otu.abun %>% 
  left_join(fun.tax) %>% 
  group_by(Phylum) %>% 
  summarise(sum=sum(abun)) %>% 
  mutate(total=sum(sum)) %>% 
  mutate(per = (sum / total)*100)

# Pinus taeda
# calculate total sequence reads per treatment group
pita.fun.reads <- fun.otu.abun %>% 
  filter(species=="PITA") %>%
  group_by(inoculum) %>%
  summarise(total=sum(abun))

# calculate the relative abundance of phyla in each treatment group
pita.fun.phy <- fun.otu.labeled %>% 
  arrange(pot.code) %>% 
  gather(OTU, abun, 2:1113) %>% 
  left_join(fun.tax.new, by="OTU") %>%
  left_join(fun.otu.pots, by="pot.code") %>% 
  filter(species=="PITA") %>%
  mutate(abun = as.numeric(abun)) %>% 
  group_by(Phylum, inoculum) %>%
  summarise(count=sum(abun)) %>% 
  left_join(pita.fun.reads, by="inoculum") %>% 
  mutate(rel = (count / total) *100) %>% 
  mutate(Phylum = str_sub(Phylum, 4, nchar(Phylum))) %>% 
  mutate(Phylum = replace_na(Phylum, "Unknown"))

# Morella cerifera 
# calculate total sequence reads per treatment group
moce.fun.reads <- fun.otu.abun %>% 
  filter(species=="MOCE") %>%
  group_by(inoculum) %>%
  summarise(total=sum(abun))

# calculate the relative abundance of phyla in each treatment group
moce.fun.phy <- fun.otu.labeled %>% 
  arrange(pot.code) %>% 
  gather(OTU, abun, 2:1113) %>% 
  left_join(fun.tax.new, by="OTU") %>%
  left_join(fun.otu.pots, by="pot.code") %>% 
  filter(species=="MOCE") %>%
  mutate(abun = as.numeric(abun)) %>% 
  group_by(Phylum, inoculum) %>%
  summarise(count=sum(abun)) %>% 
  left_join(moce.fun.reads, by="inoculum") %>% 
  mutate(rel = (count / total) *100) %>% 
  mutate(Phylum = str_sub(Phylum, 4, nchar(Phylum))) %>% 
  mutate(Phylum = replace_na(Phylum, "Unknown"))

# Ilex glabra 
# calculate total sequence reads per treatment group
ilgl.fun.reads <- fun.otu.abun %>% 
  filter(species=="ILGL") %>%
  group_by(inoculum) %>%
  summarise(total=sum(abun))

# calculate the relative abundance of phyla in each treatment group
ilgl.fun.phy <- fun.otu.labeled %>% 
  arrange(pot.code) %>% 
  gather(OTU, abun, 2:1113) %>% 
  left_join(fun.tax.new, by="OTU") %>%
  left_join(fun.otu.pots, by="pot.code") %>% 
  filter(species=="ILGL") %>%
  mutate(abun = as.numeric(abun)) %>% 
  group_by(Phylum, inoculum) %>%
  summarise(count=sum(abun)) %>% 
  left_join(ilgl.fun.reads, by="inoculum") %>% 
  mutate(rel = (count / total) *100) %>% 
  mutate(Phylum = str_sub(Phylum, 4, nchar(Phylum))) %>% 
  mutate(Phylum = replace_na(Phylum, "Unknown"))

# create df of only sterile treatment data
fun.phy.ster <- pita.fun.phy %>% 
  dplyr::select(Phylum, inoculum, rel) %>% 
  filter(inoculum=="Sterile") %>% 
  mutate(species="PITA") %>% 
  rbind(ilgl.fun.phy %>% 
              dplyr::select(Phylum, inoculum, rel) %>% 
              filter(inoculum=="Sterile") %>% 
              mutate(species="ILGL")) %>% 
  rbind(moce.fun.phy %>% 
              dplyr::select(Phylum, inoculum, rel) %>% 
              filter(inoculum=="Sterile") %>% 
              mutate(species="MOCE"))
```

Next, build data frames for each species without sterile ASVs to summarize the relative abundance of each Phylum only sourced from inoculation. 

```{r}
# create data frame of treatment data and ASV abundance
fun.otu.abun.ster <- fun.otu.inoc %>%
  gather(OTU, abun, 2:741) %>% 
  left_join(fun.tax.new, by="OTU") %>%
  left_join(fun.otu.pots, by="pot.code") %>% 
  dplyr::select(pot.code, species, inoculum, OTU, abun)

# Pinus taeda
# calculate total sequence reads per treatment group
pita.fun.reads.ster <- fun.otu.abun.ster %>% 
  filter(species=="PITA") %>%
  group_by(inoculum) %>%
  summarise(total=sum(abun))

# calculate the relative abundance of phyla in each treatment group
pita.fun.phy.ster <- fun.otu.inoc %>% 
  arrange(pot.code) %>% 
  gather(OTU, abun, 2:741) %>% 
  left_join(fun.tax.new, by="OTU") %>%
  left_join(fun.otu.pots, by="pot.code") %>% 
  filter(species=="PITA") %>%
  mutate(abun = as.numeric(abun)) %>% 
  group_by(Phylum, inoculum) %>%
  summarise(count=sum(abun)) %>% 
  left_join(pita.fun.reads.ster, by="inoculum") %>% 
  mutate(rel = (count / total) *100) %>% 
  mutate(Phylum = str_sub(Phylum, 4, nchar(Phylum))) %>% 
  mutate(Phylum = replace_na(Phylum, "Unknown"))

# Morella cerifera
# calculate total sequence reads per treatment group
moce.fun.reads.ster <- fun.otu.abun.ster %>% 
  filter(species=="MOCE") %>%
  group_by(inoculum) %>%
  summarise(total=sum(abun))

# calculate the relative abundance of phyla in each treatment group
moce.fun.phy.ster <- fun.otu.inoc %>% 
  arrange(pot.code) %>% 
  gather(OTU, abun, 2:741) %>% 
  left_join(fun.tax.new, by="OTU") %>%
  left_join(fun.otu.pots, by="pot.code") %>% 
  filter(species=="MOCE") %>%
  mutate(abun = as.numeric(abun)) %>% 
  group_by(Phylum, inoculum) %>%
  summarise(count=sum(abun)) %>% 
  left_join(moce.fun.reads.ster, by="inoculum") %>% 
  mutate(rel = (count / total) *100) %>% 
  mutate(Phylum = str_sub(Phylum, 4, nchar(Phylum))) %>% 
  mutate(Phylum = replace_na(Phylum, "Unknown"))

# Ilex glabra 
# calculate total sequence reads per treatment group
ilgl.fun.reads.ster <- fun.otu.abun.ster %>% 
  filter(species=="ILGL") %>%
  group_by(inoculum) %>%
  summarise(total=sum(abun))

# calculate the relative abundance of phyla in each treatment group
ilgl.fun.phy.ster <- fun.otu.inoc %>% 
  arrange(pot.code) %>% 
  gather(OTU, abun, 2:741) %>% 
  left_join(fun.tax.new, by="OTU") %>%
  left_join(fun.otu.pots, by="pot.code") %>% 
  filter(species=="ILGL") %>%
  mutate(abun = as.numeric(abun)) %>% 
  group_by(Phylum, inoculum) %>%
  summarise(count=sum(abun)) %>% 
  left_join(ilgl.fun.reads.ster, by="inoculum") %>% 
  mutate(rel = (count / total) *100) %>% 
  mutate(Phylum = str_sub(Phylum, 4, nchar(Phylum))) %>% 
  mutate(Phylum = replace_na(Phylum, "Unknown"))
```

### Create and assemble Phylum panels (Figure 4).

```{r}
# create color palette
phylum.colors <- c("Ascomycota"="#d40853", 
                  "Basidiomycota"="#e680a5", 
                  "Chytridiomycota"="#ffc107", 
                  "Glomeromycota"="#fbe299",
                  "Monoblepharomycota"="#0e41d0", 
                  "Mortierellomycota"="#829fef", 
                  "Mucoromycota"="#004d40", 
                  "Rozellomycota"="#81e4d3",
                  "Unknown"="gray75")

# pinus taeda figure
pita.phy.fig.ster <- ggplot(pita.fun.phy.ster) + 
  geom_bar(aes(x=inoculum, y=rel, group=Phylum, fill=Phylum), stat = "identity") + 
  theme_classic() + 
  ylab("Relative Abundance") +
  xlab("Inoculum") + 
  scale_fill_manual(values=phylum.colors) + 
  theme(plot.title = element_text(face = "italic"),
        axis.title.y = element_blank(), 
        legend.position="NONE") +  
  ggtitle("Pinus taeda")

# morella cerifera figure
moce.phy.fig.ster <- ggplot(moce.fun.phy.ster) + 
  geom_bar(aes(x=inoculum, y=rel, group=Phylum, fill=Phylum), stat = "identity") + 
  theme_classic() + 
  xlab("Inoculum") + 
  scale_fill_manual(values=phylum.colors) + 
  theme(plot.title = element_text(face = "italic"),
        axis.title.y = element_blank(), 
        legend.position="NONE") + 
  ggtitle("Morella cerifera")

# ilex glabra figure
ilgl.phy.fig.ster <- ggplot(ilgl.fun.phy.ster) + 
  geom_bar(aes(x=inoculum, y=rel, group=Phylum, fill=Phylum), stat = "identity") + 
  theme_classic() + 
  xlab("Inoculum") + 
  scale_fill_manual(values=phylum.colors) + 
  theme(plot.title = element_text(face = "italic"),
        axis.title.y = element_blank(), 
        legend.position="NONE") + 
  ggtitle("Ilex glabra")

# extract legend
fun.phy.leg.ster <- get_legend(ggplot(ilgl.fun.phy.ster) + 
  geom_bar(aes(x=inoculum, y=rel, group=Phylum, fill=Phylum), stat = "identity") + 
  theme_classic() + 
  xlab("Inoculum") + 
    scale_fill_manual(values=phylum.colors) + 
  theme(plot.title = element_text(face = "italic"),
        axis.title.y = element_blank()))

# create figure of sterile pots 
ster.pots <- ggplot(fun.phy.ster) + 
  geom_bar(aes(x=species, y=rel, group=Phylum, fill=Phylum), stat = "identity") + 
  theme_classic() + 
  ylab("Relative Abundance") +
  xlab("Species") + 
  scale_fill_manual(values=phylum.colors) + 
  theme(plot.title = element_text(face = "italic"),
        axis.title.y = element_blank(), 
        legend.position="NONE") + 
  ggtitle(" ")

# assemble first row
row1 <- ggarrange(ilgl.phy.fig.ster, moce.phy.fig.ster, pita.phy.fig.ster, nrow=1, labels=c("A", "B", "C"))
 
# assemble second row
row2 <- ggarrange(ster.pots, fun.phy.leg.ster, widths=c(2,1), nrow=1)

# combine rows with spacer in between
phyla.fig <- ggarrange(row1, row2, labels=c("", "D"), nrow=2)
phyla.fig

# export as .png for manual annotation
ggsave(plot=phyla.fig, "phyla.png", dpi=600, width=8, height=7)
```

### Most abundant classes panel (Figure 4)

```{r}
fam.colors <- c("Mortierellomycetes"="#829fef", 
           "Cystobasidiomycetes"="#e680a5", 
           "Sordariomycetes"="#d40853", 
           "Pezizomycetes"="#d40853", 
           "Tremellomycetes"="#e680a5", 
           "Saccharomycetes"="#d40853", 
           "Leotiomycetes"="#d40853", 
           "Agaricomycetes"="#e680a5", 
           "Microbotryomycetes"="#e680a5",  
           "Dothideomycetes"="#d40853")

# Most abundant classes overall 
fun.class.top <- fun.otu.labeled %>% 
  arrange(pot.code) %>% 
  gather(OTU, abun, 2:1113) %>% 
  left_join(fun.tax.new, by="OTU") %>%
  dplyr::select(Class, abun) %>% 
  group_by(Class) %>% 
  summarise(sum=sum(abun))

# Total reads per treatment group
group.reads <- fun.otu.labeled %>% 
  arrange(pot.code) %>% 
  gather(OTU, abun, 2:1113) %>% 
  left_join(fun.otu.pots, by="pot.code") %>% 
  group_by(species, inoculum) %>% 
  summarise(total.sum=sum(abun))

# Compute relative abundance for top 8 most abundant 
fun.class <- fun.otu.labeled %>% 
  arrange(pot.code) %>% 
  gather(OTU, abun, 2:1113) %>% 
  left_join(fun.tax.new, by="OTU") %>%
  dplyr::select(pot.code, Class, abun) %>% 
  left_join(fun.otu.pots, by="pot.code") %>% 
  filter(Class=="c__Mortierellomycetes" | 
           Class=="c__Cystobasidiomycetes" | 
           Class=="c__Sordariomycetes" | 
           Class=="c__Pezizomycetes" | 
           Class=="c__Tremellomycetes" | 
           Class=="c__Saccharomycetes" | 
           Class=="c__Leotiomycetes" | 
           Class=="c__Agaricomycetes" | 
           Class=="c__Microbotryomycetes" | 
           Class=="c__Dothideomycetes") %>% 
  mutate(Class = str_sub(Class, 4, nchar(Class))) %>% 
  group_by(species, inoculum, Class) %>%
  summarise(sum=sum(abun)) %>% 
  left_join(group.reads, by=c("species", "inoculum")) %>%  
  mutate(rel.abun = (sum / total.sum))

fun.class$Class <- factor(fun.class$Class, levels = c("Dothideomycetes", "Leotiomycetes", 
                                                      "Pezizomycetes", "Saccharomycetes", 
                                                      "Sordariomycetes", "Agaricomycetes", 
                                                      "Cystobasidiomycetes", 
                                                      "Microbotryomycetes", "Tremellomycetes", 
                                                      "Mortierellomycetes"))

class.plot <- ggplot(fun.class) + geom_col(aes(x=interaction(inoculum, species), y=rel.abun*100, fill=Class)) + 
  scale_fill_manual(values=fam.colors) + 
  ylab("") + 
  theme_classic() + 
  theme(legend.position = "NONE", 
        axis.text.x = element_blank(), 
        axis.title.x = element_blank(), 
        panel.background = element_rect(fill = "transparent", color = NA), 
        plot.background = element_rect(fill = "transparent", color = NA),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank()) + 
  facet_wrap(~Class, ncol=1, scales="free")
class.plot

ggsave(plot=class.plot, "myco.class.png", dpi=600, height=10, width=2.5)
```


### Ectomycorrhizal colonization via visual scoring panels  (Figure 5)

```{r}
# quantify percent colonization
myco$per.col = (myco$col / (myco$col + myco$unc))*100

# average colonization per pot
myco.per <- myco %>% 
  group_by(pot.code) %>% 
  summarise(avg.col=mean(percent.col, na.rm=TRUE)) %>% 
  left_join(pp) %>% 
  right_join(fun.guild.ecm %>% 
              dplyr::select(pot.code, rel.count), by="pot.code")

# test relationship with relative abundance of ecto sequence reads
myco.seq <- lm(rel.count ~ avg.col, data=myco.per)
summary(myco.seq)

# test relationship with plant performance
col.pp <- lm(psf~avg.col, data=myco.per)
summary(col.pp)
```

### Mycorrhizae sequencing panels + Assemble (Figure 5)

```{r}
# ECM sequence read abundance across treatments
ecm.seq.treat <- ggplot(myco.per) + 
  geom_boxplot(aes(x=inoculum, y=rel.count*100, fill=inoculum)) + 
  scale_fill_manual(values=psf.bw) + 
  ylab("Seq Read Abun") + 
  xlab("Inoculum") + 
  theme_classic() + 
  ggtitle(" ")

# ECM colonization across treatments
ecm.col.treat <- ggplot(myco.per) + 
  geom_boxplot(aes(x=inoculum, y=avg.col, fill=inoculum)) + 
  scale_fill_manual(values=psf.bw) + 
  ylab("Scored % Col") + 
  xlab("Inoculum") + 
  theme_classic() + 
  ggtitle(" ")

# relationship between ECM sequence reads and visually-scored colonization
col.seq <- ggplot(myco.per, aes(x=avg.col, y=rel.count*100)) + 
  geom_point(color="black", size=3) + 
  annotate("text", label="R2=0.4794", x=50, y=80) + 
  geom_smooth(method="lm", se=FALSE, color="black") + 
  ylim(0,100) + 
  ylab("Seq Read Abun") + xlab("Scored % Col") + 
  theme_classic() + 
  ggtitle(" ")

# relationship between ECM abundance and plant performance
pp.ecm <- ggplot(myco.per) + 
  geom_smooth(aes(x=rel.count*100, y=psf, color="Seq Read Abun"), 
              method="lm", alpha=.25, linetype="dashed") + 
  geom_smooth(aes(x=avg.col, y=psf, color="Scored % Col"), 
  method="lm", alpha=.25, linetype="dashed") + 
  geom_point(aes(x=rel.count*100, y=psf, color="Seq Read Abun"), size=3) + 
  geom_point(aes(x=avg.col, y=psf, color="% Col"), size=3) + 
  theme_classic() + 
  ylab("Rel Performance") + 
  theme(axis.title.x=element_blank()) + 
  scale_color_manual(
        name = "Variable", 
        values = c("Scored % Col" = "gray70", "Seq Read Abun" = "black"),
        labels = c("Scored % Col", "Seq Read Abun")) + 
  ggtitle("") + 
  annotate("text", label="P=0.8693", x=80, y=0.9, color="gray70") + 
  annotate("text", label="P=0.4214", x=80, y=0.7, color="black")

# AM abundance across treatments
am.seq.treat <- ggplot(fun.guild.am %>% filter(species=="ILGL")) + 
  geom_boxplot(aes(x=inoculum, y=rel.count*100, fill=inoculum)) + 
  scale_fill_manual(values=psf.bw) + 
  ylab("Seq Read Abun") + 
  xlab("Inoculum") + 
  theme_classic() + 
  ggtitle(" ")

# relationship between AM abundance and plant performance
pp.am <- ggplot(fun.guild.am %>% filter(species=="ILGL")) + 
  geom_smooth(aes(x=rel.count*100, y=psf),  
             color="black", method="lm", alpha=.25, linetype="solid") + 
  geom_point(aes(x=rel.count*100, y=psf), color="black", size=3) + 
  theme_classic() + 
  ylab("Rel Performance") + 
  theme(axis.title.x=element_blank()) + 
  ggtitle("") + 
  annotate("text", label="P=0.0065", x=13, y=1.7, color="black")

# assemble into one figure
myco.row1 <- ggarrange(am.seq.treat, ecm.seq.treat, ecm.col.treat, col.seq, 
                       common.legend=TRUE, legend="bottom", nrow=1, 
                       labels=c("A", "B", "C", "D"))

myco.row2 <- ggarrange(pp.am, pp.ecm, 
                       common.legend=TRUE, legend="bottom", labels=c("E","F"))

myco.fig <- ggarrange(myco.row1, myco.row2, nrow=2)
myco.fig

# save and annotate headings outside of R
ggsave(plot=myco.fig, "myco.figure.png", dpi=600, width=7, height=5)
```


### Random Forest Models

Prepare data 

```{r}
# filter out ASVs with abundance <0.001 
# determine what value represents relative abundance of 0.001  
sum(fun.guild.sum$total.count) # sum of all seq reads = 3696664  
0.001 * 4194981 # 0.0001% of total seq reads = 369.6664

# select only ASVs with greater than 0.0001% (rounding to the nearest whole #)
fun.rf <- fun.otu.matrix[colSums(fun.otu.matrix)>4195] 
# 127 ASVs remain

# create AVS table of selected ASVs with pot code labels
fun.rf.otu <- fun.otu.pots %>% 
  dplyr::select(pot.code) %>% 
  arrange(pot.code) %>% 
  cbind(fun.rf)
```

Run model for each species. First, Pinus taeda. 

```{r}
# create metadata df with only PITA 
pita.metadata.rf <- fun.otu.pots %>% 
  dplyr::select(pot.code) %>%
  left_join(pp, by="pot.code") %>% 
  dplyr::select(pot.code, species, inoculum, psf, bm.rate) %>% 
  filter(species=="PITA") %>% 
  arrange(pot.code)

# create otu table with only PITA
pita.fun.rf.otu <- pita.metadata.rf %>% 
  dplyr::select(pot.code) %>% 
  left_join(fun.rf.otu, by="pot.code") %>% 
  arrange(pot.code) %>%
  dplyr::select(-c(1))

# combine psf (response) and ASV table
pita.fun.rf.all <- pita.metadata.rf %>% 
  dplyr::select(psf) %>% 
  cbind(pita.fun.rf.otu) %>% 
  drop_na(psf)


# build random forest model #
# multiple iterations of model training using a "for" loop 
# set number of iterations
n_it <- 100

# create lists to store results
pita.fun.results <- list()
pita.fun.importance.list <- list()

for (i in 1:n_it) {
  set.seed(i) # set a different seed for each iteration
  
  # split data into training and testing sets
  pita.fun.train <- sample(nrow(sample_data(pita.fun.rf.all)), size = nrow(sample_data(pita.fun.rf.all))*0.7)
  pita.fun.training <- pita.fun.rf.all[pita.fun.train,]
  pita.fun.testing <- pita.fun.rf.all[-pita.fun.train,]
  
  # train the model
  pita.fun.rf.model3 <- randomForest(psf ~ ., data = pita.fun.training, importance=TRUE, ntree = 500, mtry = sqrt(ncol(pita.fun.training) - 1))
  
  # make predictions on the test set
  predictions <- predict(pita.fun.rf.model3, newdata = pita.fun.testing)
  
  # evaluate performance
  # store model output
  pita.fun.results[[i]] <- pita.fun.rf.model3
  
  # extract variable importance
  # type=2 for IncNodePurity
  pita.fun.importance <- importance(pita.fun.rf.model3, type=2)
  
  # convert to list
  pita.fun.importance.list[[i]] <- pita.fun.importance

} # consider weighting importance by model performance 

# extract importance values from each list and combine into single matrix
# each column is a model iteration
pita.fun.importance.df <- do.call(cbind, lapply(pita.fun.importance.list, function(x) x[, 1]))

# calculate correlation of ASVs with psf values 
# to use for filtering out low-correlations ASVs
pita.fun.cor <- pita.fun.rf.all %>% 
  drop_na(psf) %>% 
  correlate() %>% 
  focus(psf) %>% 
  rename("OTU"="term") %>%
  rename("corr"="psf")

# sanity check 
cor(pita.fun.rf.all$psf, pita.fun.rf.all$OTU0001)

# calculate mean importance values for each ASV
pita.fun.imp.sum <-pita.fun.importance.df %>% 
  as.data.frame() %>% 
  rownames_to_column(var="OTU") %>% 
  pivot_longer(cols=c(2:101), names_to="Iteration", values_to="Importance") %>% 
  group_by(OTU) %>% 
  summarise(Mean.Importance=mean(Importance)) %>% 
  left_join(fun.guild.gh, by="OTU") %>% 
  left_join(fun.tax, by="OTU") %>% 
  filter(Mean.Importance>0) %>% # filter out any variables (OTUs) with 0 importance
  left_join(pita.fun.cor, by="OTU") %>% 
  filter(corr>=0.30 | corr<=-0.30) # filter out anything under 0.30 R2

# are importance and correlation coefficient related?
ggplot(pita.fun.imp.sum) + 
  geom_point(aes(x=abs(corr), y=Mean.Importance)) + 
  theme_classic() + 
  ylab("Mean Importance") + 
  xlab("Absolute Correlation")

# determine 80% quantile cut off (to select only the top 20% most important)
pita.quant <- quantile(pita.fun.imp.sum$Mean.Importance, probs=0.80) 
pita.quant

# filter out lower 80%
pita.fun.imp.80 <- pita.fun.imp.sum %>% 
  filter(Mean.Importance>pita.quant)

# sanity check
# create df with individual pot reads for all important ASVs
pita.fun.imp.80.long <- pita.metadata.rf %>% 
  drop_na(psf) %>% 
  cbind(pita.fun.rf.all) %>% 
  pivot_longer(cols=c(6:128), names_to="OTU", values_to="reads") %>% 
  filter(OTU %in% pita.fun.imp.80$OTU)

# visually inspect data for quality control
# plot important ASVs and psf values 
# identify patterns driven by only 1-2 pots
ggplot(pita.fun.imp.80.long, aes(x=reads, y=psf)) + 
  geom_point(aes(color=inoculum)) + 
  geom_smooth(method="lm") + 
  facet_wrap(vars(OTU), scales="free")

# visualize data distribution by inoculum 
ggplot(pita.fun.imp.80.long, aes(x=inoculum, y=reads)) + 
  geom_boxplot(aes(color=inoculum)) + 
  facet_wrap(vars(OTU), scales="free")
```

Morella cerifera 

```{r}
# create metadata df with only MOCE #
moce.metadata.rf <- fun.otu.pots %>% 
  dplyr::select(pot.code) %>%
  left_join(pp, by="pot.code") %>% 
  dplyr::select(pot.code, species, inoculum, psf, bm.rate) %>% 
  filter(species=="MOCE") %>% 
  arrange(pot.code)

# create otu table with only PITA
moce.fun.rf.otu <- moce.metadata.rf %>% 
  dplyr::select(pot.code) %>% 
  left_join(fun.rf.otu) %>% 
  arrange(pot.code) %>%
  dplyr::select(-c(1))

# combine psf (response) and ASV table
moce.fun.rf.all <- moce.metadata.rf %>% 
  dplyr::select(psf) %>% 
  cbind(moce.fun.rf.otu) %>% 
  drop_na(psf)

# set number of iterations
n_it <- 100

# create lists to store results
moce.fun.results <- list()
moce.fun.importance.list <- list()

for (i in 1:n_it) {
  set.seed(i) # set a different seed for each iteration
  
  # split data into training and testing sets
  moce.fun.train <- sample(nrow(sample_data(moce.fun.rf.all)), size = nrow(sample_data(moce.fun.rf.all))*0.7)
  moce.fun.training <- moce.fun.rf.all[moce.fun.train,]
  moce.fun.testing <- moce.fun.rf.all[-moce.fun.train,]
  
  # train the model
  moce.fun.rf.model3 <- randomForest(psf ~ ., data = moce.fun.training, importance=TRUE, ntree = 500, mtry = sqrt(ncol(moce.fun.training) - 1))
  
  # make predictions on the test set
  predictions <- predict(moce.fun.rf.model3, newdata = moce.fun.testing)
  
  # evaluate performance
  # store model output
  moce.fun.results[[i]] <- moce.fun.rf.model3 # Or performance metrics
  
  # extract variable importance
  # type=2 for IncNodePurity
  moce.fun.importance <- importance(moce.fun.rf.model3, type=2)
  
  # convert to list
  moce.fun.importance.list[[i]] <- moce.fun.importance

}

# extract importance values from each list and combine into single matrix
# each column is a model iteration
moce.fun.importance.df <- do.call(cbind, lapply(moce.fun.importance.list, function(x) x[, 1]))

# calculate correlation of ASVs with psf values 
# to use for filtering out low-correlations ASVs
moce.fun.cor <- moce.fun.rf.all %>% 
  drop_na(psf) %>% 
  correlate() %>% 
  focus(psf) %>% 
  rename("OTU"="term") %>%
  rename("corr"="psf")

# sanity check 
cor(moce.fun.rf.all$psf, moce.fun.rf.all$OTU0001) # matches (0.010288315)

# Calculate mean importance values for each OTU and filter
moce.fun.imp.sum <-moce.fun.importance.df %>% 
  as.data.frame() %>% 
  rownames_to_column(var="OTU") %>% 
  pivot_longer(cols=c(2:101), names_to="Iteration", values_to="Importance") %>% 
  group_by(OTU) %>% 
  summarise(Mean.Importance=mean(Importance)) %>% 
  left_join(fun.guild.gh, by="OTU") %>% 
  left_join(fun.tax, by="OTU") %>% 
  filter(Mean.Importance>0) %>% 
  left_join(moce.fun.cor, by="OTU") %>% 
  filter(corr>=0.30 | corr<=-0.30)

# are importance and correlation coefficient related?
ggplot(moce.fun.imp.sum) + 
  geom_point(aes(x=abs(corr), y=Mean.Importance)) + 
  theme_classic() + 
  ylab("Mean Importance") + 
  xlab("Absolute Correlation")

# determine 80% quantile cut off (to select only the top 20% most important)
moce.quant <- quantile(moce.fun.imp.sum$Mean.Importance, probs=0.80) 
moce.quant

# filter out lower 80% 
moce.fun.imp.80 <- moce.fun.imp.sum %>% 
  filter(Mean.Importance>moce.quant)

# sanity check
# create df with individual pot reads for all important ASVs
moce.fun.imp.80.long <- moce.metadata.rf %>% 
  drop_na(psf) %>% 
  cbind(moce.fun.rf.all) %>% 
  pivot_longer(cols=c(6:128), names_to="OTU", values_to="reads") %>% 
  filter(OTU %in% moce.fun.imp.80$OTU)

# visually inspect data for quality control
# plot important ASVs and psf values 
# identify any patterns driven by only 1-2 pots
ggplot(moce.fun.imp.80.long, aes(x=log(reads), y=psf)) + 
  geom_point(aes(color=inoculum)) + 
  geom_smooth(method="lm") + 
  facet_wrap(vars(OTU), scales="free")

# visualize data distribution by inoculum
ggplot(moce.fun.imp.80.long, aes(x=inoculum, y=reads)) + 
  geom_boxplot(aes(color=inoculum)) + 
  facet_wrap(vars(OTU), scales="free")
```

Ilex glabra

```{r}
# create metadata df with only ILGL #
ilgl.metadata.rf <- fun.otu.pots %>% 
  dplyr::select(pot.code) %>%
  left_join(pp, by="pot.code") %>% 
  dplyr::select(pot.code, species, inoculum, psf, bm.rate) %>% 
  filter(species=="ILGL") %>% 
  arrange(pot.code)

# create otu table with only ILGL
ilgl.fun.rf.otu <- ilgl.metadata.rf %>% 
  dplyr::select(pot.code) %>% 
  left_join(fun.rf.otu) %>% 
  arrange(pot.code) %>%
  dplyr::select(-c(1))

# combine psf (response) and ASV table
ilgl.fun.rf.all <- ilgl.metadata.rf %>% 
  dplyr::select(psf) %>% 
  cbind(ilgl.fun.rf.otu) %>% 
  drop_na(psf)

# set number of iterations
n_it <- 100

# create lists to store results
ilgl.fun.results <- list()
ilgl.fun.importance.list <- list()

for (i in 1:n_it) {
  set.seed(i) # set a different seed for each iteration
  
  # split data into training and testing sets
  ilgl.fun.train <- sample(nrow(sample_data(ilgl.fun.rf.all)), size = nrow(sample_data(ilgl.fun.rf.all))*0.7)
  ilgl.fun.training <- ilgl.fun.rf.all[ilgl.fun.train,]
  ilgl.fun.testing <- ilgl.fun.rf.all[-ilgl.fun.train,]
  
  # train the model
  ilgl.fun.rf.model3 <- randomForest(psf ~ ., data = ilgl.fun.training, importance=TRUE, ntree = 500, mtry = sqrt(ncol(ilgl.fun.training) - 1))
  
  # make predictions on the test set
  predictions <- predict(ilgl.fun.rf.model3, newdata = ilgl.fun.testing)
  
  # evaluate performance
  # store model output
  ilgl.fun.results[[i]] <- ilgl.fun.rf.model3 # Or performance metrics
  
  # extract variable importance
  # type=2 for IncNodePurity
  ilgl.fun.importance <- importance(ilgl.fun.rf.model3, type=2)
  
  # convert to lists
  ilgl.fun.importance.list[[i]] <- ilgl.fun.importance
  
}

# extract importance values from each list and combine into single matrix
# each column is a model iteration
ilgl.fun.importance.df <- do.call(cbind, lapply(ilgl.fun.importance.list, function(x) x[, 1]))

# calculate correlation of ASVs with psf values 
# to use for filtering out low-correlations ASVs
ilgl.fun.cor <- ilgl.fun.rf.all %>% 
  drop_na(psf) %>% 
  correlate() %>% 
  focus(psf) %>% 
  rename("OTU"="term") %>%
  rename("corr"="psf")

# sanity check 
cor(ilgl.fun.rf.all$psf, ilgl.fun.rf.all$OTU0001)

# calculate mean importance values for each ASV
# filter for importance >0 and correlation > +/-0.30
ilgl.fun.imp.sum <-ilgl.fun.importance.df %>% 
  as.data.frame() %>% 
  rownames_to_column(var="OTU") %>% 
  pivot_longer(cols=c(2:101), names_to="Iteration", values_to="Importance") %>% 
  group_by(OTU) %>% 
  summarise(Mean.Importance=mean(Importance)) %>% 
  left_join(fun.guild.gh, by="OTU") %>% 
  left_join(fun.tax, by="OTU") %>% 
  filter(Mean.Importance>0) %>% 
  left_join(ilgl.fun.cor, by="OTU") %>% 
  filter(corr>=0.30 | corr<=-0.30)

# are importance and correlation coefficient related?
ggplot(ilgl.fun.imp.sum) + 
  geom_point(aes(x=abs(corr), y=Mean.Importance)) + 
  theme_classic() + 
  ylab("Mean Importance") + 
  xlab("Absolute Correlation")

# determine 80% quantile cut off (to select only the top 20% most important)
ilgl.quant <- quantile(ilgl.fun.imp.sum$Mean.Importance, probs=0.80) 
ilgl.quant

# filter out lower 80% 
ilgl.fun.imp.80 <- ilgl.fun.imp.sum %>% 
  filter(Mean.Importance>ilgl.quant)

# sanity check
# create df with individual pot reads for all important ASVs
ilgl.fun.imp.80.long <- ilgl.metadata.rf %>% 
  drop_na(psf) %>% 
  cbind(ilgl.fun.rf.all) %>% 
  pivot_longer(cols=c(6:128), names_to="OTU", values_to="reads") %>% 
  filter(OTU %in% ilgl.fun.imp.80$OTU)

# visually inspect data for quality control
# plot important ASVs and psf values 
# identify any patterns driven by only 1-2 pots
ggplot(ilgl.fun.imp.80.long, aes(x=reads, y=psf)) + 
  geom_point(aes(color=inoculum), size=2) + 
  geom_smooth(method="lm") + 
  theme_classic() + 
  facet_wrap(vars(OTU), scales="free")

# visualize data distribution by inoculum 
ggplot(ilgl.fun.imp.80.long, aes(x=inoculum, y=reads)) + 
  geom_boxplot(aes(color=inoculum)) + 
  facet_wrap(vars(OTU), scales="free")

# export combined file to CSV 
# combine the most important 20% of OTUs from all species into one df
fun.imp.80 <- pita.fun.imp.80 %>% 
  mutate(species="PITA") %>%
  rbind(moce.fun.imp.80 %>% mutate(species="MOCE"), 
        ilgl.fun.imp.80 %>% mutate(species="ILGL"))

# export as csv
write.csv(fun.imp.80, "fun.imp.80.csv")
```

### Random Forest Results Figure (Figure 6)

```{r}
# calculate mean species importance values
sp.mean.imp <- fun.imp.80 %>% group_by(species) %>% summarise(mean=mean(Mean.Importance))
sp.sd.imp <- fun.imp.80 %>% group_by(species) %>% summarise(sd=sd(Mean.Importance))

# summarize by genus
fun.imp.fig <- fun.imp.80 %>% 
  dplyr::select(species, Phylum, Order, Genus, Mean.Importance, corr) %>% 
  mutate(Phylum = str_sub(Phylum, 4, nchar(Phylum))) %>% 
  mutate(Order = str_sub(Order, 4, nchar(Order))) %>% 
  mutate(Genus = str_sub(Genus, 4, nchar(Genus))) %>% 
  drop_na(Genus) %>% 
  mutate(Phylum = forcats::fct_rev(Phylum)) %>% 
  mutate(Order = forcats::fct_rev(Order)) %>% 
  mutate(Genus = forcats::fct_rev(Genus)) %>% 
  left_join(sp.mean.imp, by="species") %>% 
  left_join(sp.sd.imp, by="species") %>% 
  mutate(rel.imp = ((Mean.Importance - mean) / sd) + 1.43)

# plot important ASVs and order taxonomically 
fun.imp.plot <- ggplot(fun.imp.fig, aes(x=species, y=interaction(Genus, Order, Phylum))) + 
  geom_point(aes(size=rel.imp, color=corr)) + 
  scale_color_gradient2(high = "#075AFF",
                       mid = "#FFFFCC",
                       low = "#FF0000") + 
  scale_size_continuous(breaks=c(0.5, 1, 1.5, 2))
fun.imp.plot

# save and annotate outside of R
ggsave(plot=fun.imp.plot, "Fungal Importance Figure.png", dpi=500, width=8, height=5)

## positive- and negative-growth associated "important" ASVs by species and treatment
fun.imp.80.dir <- fun.otu.pots %>% 
  dplyr::select(pot.code) %>%
  left_join(pp, by="pot.code") %>% 
  dplyr::select(pot.code, species, inoculum, psf, bm.rate) %>% 
  arrange(pot.code) %>% 
  cbind(fun.rf) %>% 
  pivot_longer(cols=c(6:127), names_to="OTU", values_to="reads") %>% 
  filter(OTU %in% fun.imp.80$OTU) %>% 
  left_join(fun.imp.80 %>% dplyr::select(OTU, Mean.Importance, corr, species), by=c("OTU","species")) %>% 
  drop_na(corr) %>% 
  mutate(dir=if_else(corr > 0, "pos", "neg")) %>% 
  dplyr::select(OTU, species, inoculum, corr, reads, dir) %>% 
  unique() %>% 
  group_by(species, inoculum, OTU, corr, dir) %>% 
  summarise(read.sum=sum(reads)) %>% 
  group_by(OTU) %>% top_n(1, read.sum)

# plot abundance of important ASVs per group
imp.dir <- ggplot(fun.imp.80.dir) + 
  geom_col(aes(x=inoculum, y=log(read.sum), group=dir, fill=dir), ) + 
  theme_classic() + 
  scale_fill_manual(values=c("#f16e3f","#295dfe"), name="Correlation", labels=c("Negative", "Positive")) + 
  ylab("Log(Read Abundance)") + 
  xlab("Treatment") + 
  facet_wrap(~ species)

imp.dir

ggsave(plot=imp.dir, "Important.Fungi.Direction.png", dpi=600, width=6.5, height=2)
```