---
title: "Plant Performance Analyses"
author: "Aeran Coughlin"
date: "`r Sys.Date()`"
output: pdf_document
---

This R Markdown file contains the plant performance analyses employed for the manuscript "Microbial assemblages of coastal forested wetland soil promote growth of expanding woody shrub" by Coughlin & Wright. 

### Load Packages and Data Files
```{r}
# packages
library(tidyverse)
library(car)
library(lubridate)
library(ggpubr)
library(rstatix)
library(betareg)

# raw data
germ <- read.csv("Data/PSF-exp-2022_germ-select.csv")
growth <- read.csv("Data/PSF-exp-2022_growth.csv")
bm <- read.csv("Data/PSF-exp-2022_biomass.csv")
germ <- read.csv("Data/PSF-exp-2022_germ-select.csv")
germ.all <- read.csv("Data/PSF-exp-2022_germ-all.csv")
pots <- read.csv("Data/PSF-exp-2022_pot-data.csv")
pots[pots=="MYCE"]<-"MOCE"
germ[germ=="MYCE"]<-"MOCE"
```

### Calculate growth and relative performance

```{r}
# get julian day of biomass harvest
x <- as.Date('2022-09-12')
yday(x)

# calculate growth time
start <- germ %>% select(pot.code, species, inoculum, select.date) 
start$germ.julian <- yday(as.Date(start$select.date))
grow <- growth %>% left_join(start, by="pot.code")
grow$growth.time <- 255 - grow$germ.julian

bm.grow.psf <- left_join(bm, grow, by="pot.code") %>% 
  left_join(pots %>% select(pot.code, treatment, tray, cell), by="pot.code") %>% 
  mutate(agb.rate=stem.dry.bm / growth.time) %>% 
  mutate(bgb.rate=root.dry.bm / growth.time) %>% 
  mutate(nod.rate=nodule.mass / growth.time) %>% 
  mutate(rs.ratio=root.dry.bm / stem.dry.bm) %>%
  mutate(bm.rate=total.bm / growth.time) %>% 
  mutate(rel.nod=nodule.mass / total.bm) %>% 
  mutate(rel.root.nod=nodule.mass / total.bm) %>% 
  mutate(inoc.type=ifelse(inoculum=="Sterile", "Uninoculated", "Inoculated")) %>% 
  group_by(species) %>% 
  drop_na(bm.rate) %>% 
  mutate(sp.avg=mean(bm.rate)) %>% 
  ungroup() %>% 
  mutate(psf=(bm.rate - sp.avg) / sp.avg)
```

### Calculate PSF index

```{r}
# PITA 
pita.psf <- bm.grow.psf %>% 
  filter(species=="PITA") %>% 
  filter(inoculum!="Sterile") %>% 
  select(pot.code, inoculum, bm.rate) %>% 
  pivot_wider(names_from="inoculum", values_from="bm.rate") %>% 
  select(-pot.code)

pita.psf.matrix <- outer(pita.psf$Forest, pita.psf$Shrub, `-`)

pita.psf.list <- pita.psf.matrix %>% 
  as.data.frame() %>% 
  pivot_longer(names_to="it", cols=c(1:12)) %>% 
  select(value) %>% 
  drop_na()

mean(pita.psf.list$value)
sd(pita.psf.list$value)

# MOCE
moce.psf <- bm.grow.psf %>% 
  filter(species=="MOCE") %>% 
  filter(inoculum!="Sterile") %>% 
  select(pot.code, inoculum, bm.rate) %>% 
  pivot_wider(names_from="inoculum", values_from="bm.rate") %>% 
  select(-pot.code)

moce.psf.matrix <- outer(moce.psf$Shrub, moce.psf$Forest, `-`)

moce.psf.list <- moce.psf.matrix %>% 
  as.data.frame() %>% 
  pivot_longer(names_to="it", cols=c(1:11)) %>% 
  select(value) %>% 
  drop_na()

mean(moce.psf.list$value)
sd(moce.psf.list$value)

# ILGL
ilgl.psf <- bm.grow.psf %>% 
  filter(species=="ILGL") %>% 
  filter(inoculum!="Sterile") %>% 
  select(pot.code, inoculum, bm.rate) %>% 
  pivot_wider(names_from="inoculum", values_from="bm.rate") %>% 
  select(-pot.code)

ilgl.psf.matrix <- outer(ilgl.psf$Shrub, ilgl.psf$Forest, `-`)

ilgl.psf.list <- ilgl.psf.matrix %>% 
  as.data.frame() %>% 
  pivot_longer(names_to="it", cols=c(1:10)) %>% 
  select(value) %>% 
  drop_na()

mean(ilgl.psf.list$value)
sd(ilgl.psf.list$value)
```

### Quantify germination 

```{r}
# create germination yes/no binary variable
germ.yn <- germ.all %>% 
  mutate(y.n=ifelse(germ.day>0, 1, y.n)) %>% 
  mutate(y.n = replace_na(y.n, 0))

# calculate percent germination and analyze inoculum effect
germ.per <- germ %>% select(pot.code,species,inoculum, seeds.planted, seeds.germ)
germ.per$percent.germ <- (germ.per$seeds.germ / germ.per$seeds.planted)
germ.per.summ <- germ.per %>% group_by(species,inoculum) %>% summarise(avg=mean(percent.germ))

# par down df an convert dates to julian day
germ.time <- germ.all %>% 
  select(pot.code, species, inoculum, plant.date, germ.date) %>% 
  drop_na(germ.date)
germ.time$plant.julian <- yday(as.Date(germ.time$plant.date))
germ.time$germ.julian <- yday(as.Date(germ.time$germ.date))

# calculate days to germination from planting
germ.time$days.to.germ <- germ.time$germ.julian - germ.time$plant.julian

# average days to germination by group
germ.time.summ <- germ.time %>% 
  drop_na(germ.julian) %>% 
  group_by(species,inoculum) %>% 
  summarise(avg=mean(days.to.germ))
```


### Testing predictors for assumptions of normality

Source: https://www.datanovia.com/en/lessons/normality-test-in-r/#google_vignette

Shapiro Test - the p-value > 0.05 implies that the distribution of the data are 
not significantly different from normal distribution.

```{r}
# growth rate 
ggdensity(sqrt(bm.grow.psf$bm.rate), fill = "lightgray")
ggqqplot(bm.grow.psf$bm.rate)
bm.grow.psf %>%
  group_by(species) %>%
  mutate(bm.rate=sqrt(bm.rate)) %>% 
  shapiro_test(bm.rate)
# Skewed slightly positive, but only MOCE seems to be non-normal
# Log transformation makes MOCE meet assumptions of normality 

# germination rate
ggdensity(germ.per$percent.germ, fill = "lightgray")
ggqqplot(germ.per$percent.germ)
germ.per %>%
  group_by(species) %>%
  shapiro_test(percent.germ)
# Bound between 0 - 1; Not normally distributed, skewed positive; will use beta regression

# germination time
ggdensity(germ.time$days.to.germ, fill = "lightgray")
ggqqplot(germ.time$days.to.germ)
germ.time %>%
  group_by(species) %>%
  mutate(days.to.germ=log10(days.to.germ)) %>% 
  shapiro_test(days.to.germ)
# Skewed strongly positive, will possibly use GLM (Inverse Gaussian with Log Link)

# aboveground biomass
ggdensity(bm.grow.psf$agb.rate, fill = "lightgray")
ggqqplot(bm.grow.psf$agb.rate)
bm.grow.psf %>%
  mutate(agb.rate=sqrt(agb.rate)) %>% 
  group_by(species) %>%
  shapiro_test(agb.rate)

# belowground biomass
ggdensity(bm.grow.psf$bgb.rate, fill = "lightgray")
ggqqplot(bm.grow.psf$bgb.rate)
bm.grow.psf %>%
  group_by(species) %>%
  shapiro_test(bgb.rate)

# root:shoot
ggdensity(bm.grow.psf$rs.ratio, fill = "lightgray")
ggqqplot(bm.grow.psf$rs.ratio)
bm.grow.psf %>%
  group_by(species) %>%
  shapiro_test(rs.ratio)
```

### Test for differences in response variables between species * inoculum

```{r}
# growth rate
growth.lm <- lm(bm.rate ~ species * inoculum, data=bm.grow.psf)
Anova(growth.lm)
summary(growth.lm)

# germination rate
germ.rate.br <- betareg(percent.germ ~ species * inoculum, data=germ.per)
summary(germ.rate.br)
Anova(germ.rate.br)

# time to germination
ttg.glm <- glm(days.to.germ ~ species * inoculum, family=inverse.gaussian(link="log"), data=germ.time)
summary(ttg.glm)

ttg.lm <- lm(days.to.germ ~ species * inoculum, data=germ.time)
Anova(ttg.lm)
summary(ttg.lm)
qqnorm(residuals(ttg.glm, type="deviance"))

# aboveground biomass
# AIC of lm was lower that glm; glm generally not a good fit #
agb.lm <- lm(agb.rate ~ species * inoculum, data=bm.grow.psf)
summary(agb.lm)
Anova(agb.lm)

# belowground biomass
bgb.lm <- lm(bgb.rate ~ species * inoculum, data=bm.grow.psf)
summary(bgb.lm)
Anova(bgb.lm)

# root:shoot ratio
rsr.lm <- lm(rs.ratio ~ species * inoculum, data=bm.grow.psf)
summary(rsr.lm)
Anova(rsr.lm)

# mean and sd of root:shoot ratio
bm.grow.psf %>% group_by(species, inoculum) %>% drop_na(rs.ratio) %>% summarise(mean=mean(rs.ratio))
bm.grow.psf %>% group_by(species) %>% drop_na(rs.ratio) %>% summarise(sd=sd(rs.ratio))
```
### Analyze differences between treatment groups within species

```{r}
# relative performance

# PITA
pita.bm.lm <- lm(psf ~ inoculum, data=bm.grow.psf %>% filter(species=="PITA"))
summary(pita.bm.lm)

qqnorm(residuals(pita.bm.lm, type="deviance"))

pita.bm.aov <- aov(pita.bm.lm)
TukeyHSD(pita.bm.aov, conf.level=0.95)

# MOCE
moce.bm.lm <- lm(psf ~ inoculum, data=bm.grow.psf %>% filter(species=="MOCE"))
summary(moce.bm.lm)

qqnorm(residuals(moce.bm.lm, type="deviance")) # More normal

moce.bm.aov <- aov(moce.bm.lm)
TukeyHSD(moce.bm.aov, conf.level=0.95)

# ILGL
ilgl.bm.lm <- lm(psf ~ inoculum, data=bm.grow.psf %>% filter(species=="ILGL"))
summary(ilgl.bm.lm)

qqnorm(residuals(ilgl.bm.lm, type="deviance"))

ilgl.bm.aov <- aov(ilgl.bm.lm)
TukeyHSD(ilgl.bm.aov, conf.level=0.95)

# aboveground biomass

# MOCE
# AIC of lm was lower that glm; glm generally not a good fit #
agb.lm.moce <- lm(log10(agb.rate) ~ inoculum, data=bm.grow.psf %>% filter(species=="MOCE"))
summary(agb.lm.moce)
Anova(agb.lm.moce)

# ILGL
agb.lm.ilgl <- lm(agb.rate ~ inoculum, data=bm.grow.psf %>% filter(species=="ILGL"))
summary(agb.lm.ilgl)
Anova(agb.lm.ilgl)

# PITA 
agb.lm.pita <- lm(agb.rate ~ inoculum, data=bm.grow.psf %>% filter(species=="PITA"))
summary(agb.lm.pita)
Anova(agb.lm.pita)

# belowground biomass

# MOCE
bgb.lm.moce <- lm(log10(bgb.rate) ~ inoculum, data=bm.grow.psf %>% filter(species=="MOCE"))
summary(bgb.lm.moce)
Anova(bgb.lm.moce)

# ILGL
bgb.lm.ilgl <- lm(bgb.rate ~ inoculum, data=bm.grow.psf %>% filter(species=="ILGL"))
summary(bgb.lm.ilgl)
Anova(bgb.lm.ilgl)

# PITA
bgb.lm.pita <- lm(bgb.rate ~ inoculum, data=bm.grow.psf %>% filter(species=="PITA"))
summary(bgb.lm.pita)
Anova(bgb.lm.pita)

## nodule biomass rate (MOCE only)

nod.lm <- lm(nod.rate ~ inoculum, data=bm.grow.psf %>% filter(species=="MOCE"))
summary(nod.lm)
Anova(nod.lm)
```

### Analyze differences between species in inoculum reponse

```{r}
# Inoculated versus Uninoculated 
inoc.psf <- lm(psf~inoc.type, data=bm.grow.psf)
Anova(inoc.psf)
ggplot(bm.grow.psf) + geom_boxplot(aes(x=inoc.type, y=psf))

# species differences in forest soil
for.psf <- lm(psf~species, data=bm.grow.psf %>% filter(inoculum=="Forest"))
summary(for.psf)
ggplot(bm.grow.psf %>% filter(inoculum=="Forest")) + geom_boxplot(aes(x=species, y=psf))

# species differences in shrub soil
shr.psf <- lm(psf~species, data=bm.grow.psf %>% filter(inoculum=="Shrub"))
summary(shr.psf)
ggplot(bm.grow.psf %>% filter(inoculum=="Shrub")) + geom_boxplot(aes(x=species, y=psf))

# species differences in sterile soil
ste.psf <- lm(psf~species, data=bm.grow.psf %>% filter(inoculum=="Sterile"))
summary(ste.psf)
ggplot(bm.grow.psf %>% filter(inoculum=="Sterile")) + geom_boxplot(aes(x=species, y=psf))
```

# Relative performance across species and inocula (Figure 2)

```{r}
psf.bw <- c("Forest"="gray30", "Shrub"="gray", "Sterile"="gray95")

psf.values <- ggplot(bm.grow.psf %>% 
         mutate(species = replace(species, species=="ILGL", "Ilex glabra")) %>%  
       mutate(species = replace(species, species=="MOCE", "Morella cerifera")) %>%  
       mutate(species = replace(species, species=="PITA", "Pinus taeda"))) + 
  geom_hline(yintercept=0, linetype="dotted") + 
  geom_boxplot(aes(x=inoculum, y=psf, fill=inoculum)) + 
  ylab("Relative Performance") + 
  xlab("Inoculum") + 
  scale_fill_manual(values = psf.bw) + 
  theme(legend.position="NONE", 
        strip.text = element_text(face = "italic"),
        strip.background = element_rect(fill="transparent"),
        panel.background = element_rect(fill = "transparent", color = NA),
        plot.background = element_rect(fill = "transparent", color = NA),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        panel.border = element_rect(color = "black", fill = NA, linewidth = 1)) + 
  facet_wrap(~ species)
psf.values

ggsave("Figures/Figure 2 BW.png", plot=psf.values, width = 5, height = 3.5, device='tiff', dpi=500, bg = "transparent")
```
